# Analysis

```{r setup-analysis}
#| include: false
#| echo: false
#| warning: false
#| message: false

library(arrow)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggrepel)
library(ggcorrplot)
library(strucchange)
library(car)

regional  <- read_parquet("data/analysis_panel_regional.parquet")
statewide <- read_parquet("data/analysis_panel_statewide.parquet")

regional <- regional %>%
  mutate(yield_derived = production_tons / acres_harvested
         , variety_label = str_trim(variety_label)
         , variety_label = if_else(variety_label == "Riesling/White Riesling", "Riesling", variety_label))
statewide <- statewide %>%
  mutate(yield_derived = production_tons / acres_harvested
         , variety_label = str_trim(variety_label)
         , variety_label = if_else(variety_label == "Riesling/White Riesling", "Riesling", variety_label))

key_varieties <- c("Pinot Noir", "Pinot Gris", "Chardonnay", "Riesling")

variety_colors <- c(
  "Pinot Noir" = "#8B1E3F"
  , "Pinot Gris" = "#4393c3"
  , "Chardonnay" = "#d6604d"
  , "Riesling" = "#d4a017"
  , "Other" = "#999999"
)

region_levels <- regional %>%
  filter(!region_label %in% c("Eastern Oregon", "Other Region", "Other Oregon")) %>%
  distinct(region_label) %>%
  arrange(region_label) %>%
  pull(region_label)

# region_colors <- setNames(
#   RColorBrewer::brewer.pal(max(3, length(region_levels)), "Set1")[seq_along(region_levels)]
#   , region_levels
# )

region_colors <- c(
  "North Willamette Valley" = "#8B1E3F",
  "South Willamette Valley" = "#4393c3",
  "Rogue Valley"            = "#d6604d",
  "Umpqua"                  = "#d4a017",
  "Columbia River at Large" = "#999999"
)

lighten_color <- function(hex, amount = 0.65) {
  rgb_val <- grDevices::col2rgb(hex) / 255
  new_rgb <- rgb_val + (1 - rgb_val) * amount
  grDevices::rgb(new_rgb[1], new_rgb[2], new_rgb[3])
}
variety_colors_light <- setNames(sapply(variety_colors, lighten_color), names(variety_colors))
region_colors_light <- setNames(sapply(region_colors, lighten_color), names(region_colors))
```

## Oregon's growing reliance on Pinot Noir

```{r pinot_dominance}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-cap: "Figure 1: Statewide grape tonnage by variety, 1987–2024"
variety_share <- statewide %>%
  filter(!is.na(production_share_pct)
    , !is_zero_production
    , variety_label != "All"
    , yr >= 1987) %>%
  mutate(variety_group = if_else(variety_label %in% key_varieties, variety_label, "Other")) %>%
  group_by(yr, variety_group) %>%
  summarise(production_tons = sum(production_tons, na.rm = TRUE), .groups = "drop") %>%
  group_by(yr) %>%
  mutate(production_share_pct = production_tons / sum(production_tons, na.rm = TRUE) * 100) %>%
  ungroup()

labels_df <- variety_share %>%
  group_by(variety_group) %>%
  filter(yr == max(yr)) %>%
  ungroup()

variety_share %>%
  mutate(is_pinot = variety_group == "Pinot Noir") %>%
  ggplot(aes(yr, production_share_pct, color = variety_group, alpha = is_pinot, linewidth = is_pinot)) +
  geom_line(alpha = 0.2, linewidth = 0.4) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4, aes(linewidth = is_pinot)) +
  geom_text_repel(
      data = labels_df, aes(x = yr, y = production_share_pct, label = variety_group, color = variety_group)
      , nudge_x = 1
      , hjust = 0
      , size = 3.2
      , segment.color = variety_colors_light
      , segment.size = 0.3
      , direction = "y"
      , show.legend = FALSE
      , inherit.aes = FALSE
    ) +
  scale_color_manual(values = variety_colors) +
  scale_linewidth_manual(values = c(`TRUE` = 1.4, `FALSE` = 0.8)) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.7)) +
  scale_x_continuous(limits = c(1987, 2028), breaks = seq(1987, 2024, 5)) +
  labs(title = "Pinot Noir has come to define Oregon wine"
       , subtitle = "Share of statewide tonnage by variety, 1987–2024"
       , x = NULL, y = "Production share (%)") +
  theme_minimal() +
  theme(legend.position = "none"
        , plot.title = element_text(size = 16, face = "bold")
        , plot.subtitle = element_text(size = 10, color = "grey40"))
```

```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false
# Small helper to pull a variety's share for a given year
get_share <- function(df, variety, year, span = 0.4) {
  sub <- df %>% filter(variety_group == variety)
  fit <- loess(production_share_pct ~ yr, data = sub, span = span)
  predict(fit, newdata = data.frame(yr = year)) %>% round(0)
}

pinot_end_pct <- get_share(variety_share, "Pinot Noir", 2020)

pinot_start_yr <- 1990
pinot_end_yr <- 2020
gris_start_yr <- 1990
gris_end_yr <- 2010

pinot_start_pct <- get_share(variety_share, "Pinot Noir", pinot_start_yr)
pinot_end_pct <- get_share(variety_share, "Pinot Noir", pinot_end_yr)
gris_start_pct <- get_share(variety_share, "Pinot Gris", gris_start_yr)
gris_end_pct <- get_share(variety_share, "Pinot Gris", gris_end_yr)
```

Figure 1 demonstrates Pinot Noir's growing share of statewide tonnage, from roughly `r pinot_start_pct`% in `r pinot_start_yr` to around `r pinot_end_pct`% by `r pinot_end_yr`. This trend has continued to build, while other anchor varieties like Chardonnay and Riesling have declined sharply over the same period. Pinot Gris is one variety that counteracts this trend, growing from a negligible share in `r gris_start_yr` to a stable ~`r gris_end_pct`% by `r gris_end_yr`.

However, the dramatic shift toward Pinot Noir has remained a dominant trend, suggesting that Oregon's wine industry has increasingly concentrated on climate-sensitive Pinot varieties over exactly the same decades that have coincided with a pronounced shift in the region's key climate indicators.

## Evidence of a warming climate

```{r gdd_statewide}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-cap: "Figure 2: Growing degree day (GDD) anomaly relative to the 38-year average, overlaid with a 10-year rolling average"
sw_climate <- statewide %>%
  distinct(
    yr
    , gdd_apr_sep
    , gdd_apr_oct
    , frost_days_annual
    , frost_days_spring
    , heat_stress_days
    , summer_tmax_mean
    , oct_ppt_mm
    , annual_ppt_mm
    , vpd_max_summer
    , diurnal_range_ripening
    ) %>%
  arrange(yr) %>%
  mutate(
    gdd_rolling10 = zoo::rollmean(gdd_apr_sep, 10, fill = NA, align = "right")
    , gdd_anom = gdd_apr_sep - mean(gdd_apr_sep, na.rm = TRUE)
    , gdd_rolling10_anom = gdd_rolling10 - mean(gdd_apr_sep, na.rm = TRUE)
    )

# Structural-break test on the raw GDD series
gdd_ts <- ts(sw_climate$gdd_apr_sep, start = min(sw_climate$yr))
bp_fit <- breakpoints(gdd_ts ~ 1)
break_idx  <- bp_fit$breakpoints[1]
break_year <- sw_climate$yr[break_idx]

ggplot(sw_climate, aes(yr, gdd_anom)) +
  geom_col(aes(fill = gdd_anom > 0), width = 0.7) +
  geom_line(aes(y = gdd_rolling10_anom), color = "black", linewidth = 1) +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.4) +
  scale_fill_manual(values = c(`TRUE` = "#8B1E3F", `FALSE` = "#2166ac"), guide = "none") +
  geom_vline(xintercept = break_year, linetype = "dashed", color = "grey40", linewidth = 0.6) +
  annotate("text", x = break_year, y = -150, label = paste0("Structural shift begins in ", break_year), hjust = -0.15, vjust = 1, size = 3, color = "grey40") +
  labs(title = "Oregon's growing season is getting warmer"
       , subtitle = "Growing Degree Days (GDD) anomaly from the 1987–2024 mean"
       , x = NULL, y = "GDD (Apr–Sep) anomaly") +
  theme_minimal() +
  theme(legend.position = "none"
      , plot.title = element_text(size = 16, face = "bold")
      , plot.subtitle = element_text(size = 10, color = "grey40"))
```

Growing degree days (GDDs) show a clear and sustained upward trend, with a level shift identified by a structural-break test around `r break_year`. 

```{r climate_indicators}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-cap: "Figure 3: Six viticulture-relevant climate indicators, statewide average, 1987–2024, each with a linear trend and 95% confidence interval"
sw_climate %>%
  select(yr
         , `Frost risk days` = frost_days_spring
         , `GDD (Apr-Sep)` = gdd_apr_sep
         , `Peak summer temperature (°C)` = summer_tmax_mean
         , `Heat stress days` = heat_stress_days
         , `Diurnal ripening temperature range (°C)` = diurnal_range_ripening
         , `October precipitation (mm)` = oct_ppt_mm
         ) %>%
  pivot_longer(-yr, names_to = "indicator", values_to = "value") %>%
  mutate(indicator = factor(indicator, levels = c(
    "Frost risk days"
    , "GDD (Apr-Sep)"
    , "Peak summer temperature (°C)"
    , "Heat stress days"
    , "Diurnal ripening temperature range (°C)"
    , "October precipitation (mm)"
  ))) %>%
  ggplot(aes(yr, value)) +
  geom_line(color = "grey60", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "#8B1E3F", fill = "#8B1E3F", alpha = 0.15) +
  facet_wrap(~indicator, scales = "free_y", ncol = 2) +
  labs(title = str_wrap("Diurnal ripening range is a notable exception among Oregon's viticultural indicators", width = 60)
       , subtitle = "Oregon viticultrual indicators by season; linear trend lines with 95% confidence intervals, 1987-2024"
       , x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "none"
      , plot.title = element_text(size = 16, face = "bold")
      , plot.subtitle = element_text(size = 10, color = "grey40"))
```

Figure 3 summarizes six key viticultural climate indicators. Growing degree days, peak summer temperature, and heat stress days are all trending up with tight confidence intervals, demonstrating a warming signal that is both real and steep. Spring frost days are essentially flat, a meaningful non-finding since it means the warming trend has not reduced the risk of a damaging frost event during spring bud break, even as summer heat accumulation climbs. October precipitation trends slightly upward but noisily, with no strong signal either way.

The sixth indicator—ripening-season diurnal temperature range—is the one metric that moves in the opposite direction from the rest, and it does so specifically because it isn't a simple byproduct of an overall upward heat trend. This finding is most directly relevant to Oregon's growing dependence on Pinot Noir, since Pinot Noir relies on a wide day-night temperature swing during ripening to retain acidity and develop aromatic complexity—a mechanism we discussed in the [Background](#background).

## How the climate features relate

Several of these climate measures are closely related to aspects of the same underlying warming trend. To identify features that connect to the same trend, we ran a correlation matrix. 

```{r climate_correlations}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-cap: "Figure 4: Correlation matrix of annual climate features, statewide, 1987–2024"
# Aggregate to one row per year (mean across regions) before computing correlations
ry <- regional %>%
  distinct(
    region_label
    , yr
    , gdd_apr_sep
    , gdd_veraison_harvest
    , frost_days_spring
    , heat_stress_days
    , coldest_spring_night_c
    , oct_ppt_mm
    , sep_oct_ppt_mm
    , gs_ppt_mm
    , vpd_max_summer
    , diurnal_range_ripening
    )

climate_yearly <- ry %>%
  group_by(yr) %>%
  summarise(across(
    c(gdd_apr_sep, gdd_veraison_harvest, frost_days_spring, heat_stress_days
      , coldest_spring_night_c, oct_ppt_mm, sep_oct_ppt_mm, gs_ppt_mm
      , vpd_max_summer, diurnal_range_ripening)
    , mean, na.rm = TRUE
  ), .groups = "drop")

corr_matrix <- climate_yearly %>%
  select(
    frost_days_spring
    , coldest_spring_night_c
    , gdd_veraison_harvest
    , gdd_apr_sep
    , heat_stress_days
    , vpd_max_summer
    , diurnal_range_ripening
    , gs_ppt_mm
    , sep_oct_ppt_mm
    , oct_ppt_mm
  ) %>%
  rename(
    `Frost risk days` = frost_days_spring
    , `Coldest spring night` = coldest_spring_night_c
    , `GDD Veraison-Harvest` = gdd_veraison_harvest
    , `GDD Apr-Sep` = gdd_apr_sep
    , `Heat stress days` = heat_stress_days
    , `Summer VPD` = vpd_max_summer
    , `Diurnal temp range` = diurnal_range_ripening
    , `Growing season precip` = gs_ppt_mm
    , `Sep-Oct precip` = sep_oct_ppt_mm
    , `Oct precip` = oct_ppt_mm
  ) %>%
  cor(use = "pairwise.complete.obs")

ggcorrplot(
  corr_matrix
  , type = "upper"
  , lab = TRUE
  , lab_size = 3
  , colors = c("#4393c3", "white", "#8B1E3F")
  )
```

Figure 4 identifies three clusters of variables that appear to measure largely the same underlying phenomenon from different angles, rather than independent drivers:

1. **Spring frost risk:** Frost risk days and coldest spring night are correlated at -0.91, which makes sense given one is a temperature reading and the other is a threshold-count derived from it.
2. **Summer heat accumulation:** Heat stress days and summer VPD are both correlated above 0.6 with all the GDD variables.
3. **Fall precipitation:** Precipitation measures move together above 0.4.

We also note that diurnal ripening temperature range is negatively correlated with nearly every heat-related feature, consistent with the pattern described in the [Background](#background), where nights warm faster than days as the season heats up overall, compressing the day-night swing. Additionally, growing-season precipitation is negatively correlated with GDD Apr-Sep (-0.11), indicating that hotter years also tend to be drier years overall.

To resolve which variables to retain for modeling, we turn to variance inflation factors (VIF) to capture multi-variable overlap. When looking at VIF values, the lower the number, the less multicollinearity there is between features. 

```{r climate_vif}
#| echo: false
#| warning: false
#| message: false
# Setup
set.seed(42)
plot_vif <- function(vif_vals) {
  data.frame(Variable = names(vif_vals), VIF = round(unname(vif_vals), 2)) %>%
    mutate(Variable = dplyr::recode(Variable
      , "frost_days_spring" = "Frost risk days"
      , "coldest_spring_night_c" = "Coldest spring night"
      , "gdd_veraison_harvest" = "GDD Veraison-Harvest"
      , "gdd_apr_sep" = "GDD Apr-Sep"
      , "heat_stress_days" = "Heat stress days"
      , "vpd_max_summer" = "Summer VPD"
      , "diurnal_range_ripening" = "Diurnal temp range"
      , "gs_ppt_mm" = "Growing season precip"
      , "sep_oct_ppt_mm" = "Sep-Oct precip"
      , "oct_ppt_mm" = "Oct precip"
    )) %>%
    arrange(desc(VIF)) %>%
    knitr::kable(caption = "Variance Inflation Factors", row.names = FALSE)
}

ry_clean <- ry %>%
  filter(!region_label %in% c("Eastern Oregon", "Other Region", "Other Oregon"))

vif_df <- ry_clean %>%
  mutate(dummy_outcome = rnorm(n()))

# VIF - all variables
vif_model <- lm(dummy_outcome ~ gdd_apr_sep
  + frost_days_spring
  + coldest_spring_night_c
  + gdd_veraison_harvest
  + heat_stress_days
  + vpd_max_summer
  + diurnal_range_ripening
  + sep_oct_ppt_mm
  + oct_ppt_mm
  , data = vif_df
  )
vif_vals <- car::vif(vif_model)
plot_vif(vif_vals)
```

Sep-Oct precipitation and Oct precipitation come out highest, confirming we should not include both in the model. We elected to drop Sep-Oct precipitation, as we recognize Oct precipitation specifically as an important botrytis / harvest-timing risk.

Surprisingly, Summer VPD is *higher* than GDD Apr-Sep (6.43 vs. 4.13), even though the correlation matrix showed VPD's strongest pairwise correlation was with GDD. This tells us Summer VPD is well-explained by some *combination* of GDD Apr-Sep, heat stress days, and ripening temp range together, rather than by a strong individual pairwise correlation with any one of them. We elect to drop GDD Apr-Sep and keep Summer VPD, as it is the more direct measure of vine water stress.

```{r climate_vif_3}
#| echo: false
#| warning: false
#| message: false
# VIF - final features
vif_model_2 <- lm(dummy_outcome ~ gdd_veraison_harvest 
  + frost_days_spring 
  + coldest_spring_night_c
  + heat_stress_days 
  + vpd_max_summer 
  + diurnal_range_ripening
  + oct_ppt_mm 
  , data = vif_df
  )
vif_vals <- car::vif(vif_model_2)
plot_vif(vif_vals)
```

All remaining VIFs fall under a conservative threshold of VIF<5. Our retained predictor set going into the modeling stage is:

| Feature | Definition |
|-|-|
| Frost risk days | Count of days below freezing during the April–May bud-break window |
| Coldest spring night | Minimum recorded temperature during spring bud-break |
| GDD (Veraison–Harvest) | Accumulated heat above a 10°C (50°F) base during July 15–October 15 "veraison-to-harvest" window |
| Heat stress days | Count of days with a maximum temperature above 35°C (95°F) during the growing season |
| Summer VPD | A measure of atmospheric dryness / evaporative demand during July–August |
| Diurnal temperature range (ripening) | Average day-night temperature swing during August–September, the period when grapes develop their sugar, color, and acidity |
| Harvest precipitation (October) | Total precipitation in October |

Note that GDD Apr-Sep itself remains central to the *descriptive* climate trends in Figures 2–3 above; it is excluded only from the predictive feature set, where its information is substantially redundant with the other retained heat measures.